Figure two

### 2025-10-27 draftFiguresPaper.Rmd
figure 2, by exposure

datasetPollution_avg <- datasetPollution %>% 
  filter(Year == 2019) %>% 
  group_by(Region) %>% 
  summarize(pm25Avg = mean(Population.Weighted.PM2.5..ug.m3., na.rm=T))

@@

@@

## save this dataset for further analyses
write.csv(surveyCountries_long_clean, "output/datasetSurveyCountries_long_clean.csv", row.names = F)

>>>>

@@

### get dataset figure 2 papers
data_fig2_papers <- left_join(datasetClassificationCountries_long_clean, datasetPollution_ctr, by = "iso3")
write.csv(data_fig2_papers,"dataNew/data_fig2_papers.csv", row.names = FALSE)

### papers by country plus pollution
papers_by_country_pollution <- merge(papers_by_country, datasetPollution_ctr, by = "iso3", all.x = T, all.y = T)

@@

### get dataset figure 2 suveys
data_fig2_surveys <- left_join(surveyCountries_long_clean, datasetPollution_ctr, by = "iso3")
write.csv(data_fig2_surveys,"dataNew/data_fig2_surveys.csv", row.names = FALSE)

### 2025-10-27 draftFiguretwo_Ella.Rmd
figure 2, by exposure

#-------------------------------------------------------------------------------
# Load the data
#-------------------------------------------------------------------------------
papers <- read.csv2("output/datasetClassificationCountries_long_clean.csv", 
                    header = T, sep = ",", dec = ".")

surveys <- read.csv2("output/datasetSurveyCountries_long_clean.csv", 
                     header = T, sep = ",", dec = ".")

@@

#-------------------------------------------------------------------------------
# relevant columns
#-------------------------------------------------------------------------------
papers <- papers[,c("ID", "country_clean",
                    "Income.group", "Region", 
                    "Perceptions", "Behavior", "Priority", "Policy", "Health", "None")]

surveys <- surveys[,c("id", "country_clean", "year", "Income.group", "Region", 
                    "Perceptions", "Behavior", "Priority", "Policy", "Health", "Inclusion")]

@@

## Visualizations by subgroup

```{r subgroupies}

@@

# 1. Prepare topic-level data (shares within income group)
topics_data <- papers %>%
  pivot_longer(
    cols = c(Perceptions, Behavior, Priority, Policy, Health, None),
    names_to = "Topic",
    values_to = "Value") %>%
  filter(Value == 1, !is.na(Income.group)) %>%
  group_by(Topic, Income.group) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(Income.group) %>%
  mutate(share = n / sum(n),
         total = sum(n)) %>%
  ungroup() %>%
  mutate(Income.group = factor(
    Income.group,
    levels = c("Low income", "Lower middle income", "Upper middle income", "High income")))

# 2. Prepare totals row data (absolute counts)
totals_data <- topics_data %>%
  distinct(Income.group, total) %>%
  mutate(Topic = "Total papers")

@@

# 1. Prepare topic-level data (shares within income group)
topics_data_s <- surveys %>%
  pivot_longer(
    cols = c(Perceptions, Behavior, Priority, Policy, Health),
    names_to = "Topic",
    values_to = "Value") %>%
  filter(Value == 1, !is.na(Income.group)) %>%
  group_by(Topic, Income.group) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(Income.group) %>%
  mutate(share = n / sum(n),
         total = sum(n)) %>%
  ungroup() %>%
  mutate(Income.group = factor(
    Income.group,
    levels = c("Low income", "Lower middle income", "Upper middle income", "High income")))

# 2. Prepare totals row data (absolute counts)
totals_data_s <- topics_data %>%
  distinct(Income.group, total) %>%
  mutate(Topic = "Total papers")

# 3. Define consistent factor levels for the y-axis
topic_levels_s <- c("Total items", unique(topics_data$Topic))

# 4. Plot with two independent fill scales
pdf("plots/Figure2_category_survey_by_income.pdf", width = 10)
ggplot() +
  # --- top totals row ---
  geom_tile(data = totals_data_s,
            aes(x = Income.group, y = "Total items", fill = total),
            color = "white", linewidth = 0.8) +
  geom_text(
    data = totals_data_s,
    aes(x = Income.group, y = "Total items",
        label = paste0("n=", total),
        color = ifelse(total > max(topics_data_s$total) * 0.4, "black", "black")), size = 4) +
  scale_fill_gradient(
    name = "Total items",
    low = "#F6DBA7", high = "#F7B32B",
    breaks = range(totals_data_s$total),         
  labels = range(totals_data_s$total)) +
  scale_color_identity() +

@@

>>>>

#-------------------------------------------------------------------------------
# Load the data
#-------------------------------------------------------------------------------
papers <- read.csv2("dataNew/data_fig2_papers.csv", 
                    header = T, sep = ",", dec = ".")

surveys <- read.csv2("dataNew/data_fig2_surveys.csv", 
                     header = T, sep = ",", dec = ".")

@@

#-------------------------------------------------------------------------------
# relevant columns
#-------------------------------------------------------------------------------
papers <- papers[,c("ID", "country_clean",
                    "Income.group", "Region.x", 
                    "Perceptions", "Behavior", "Priority", "Policy", "Health", "None", "pm25Avg")]

surveys <- surveys[,c("id", "country_clean", "year", "Income.group", "Region.x", 
                      "Perceptions", "Behavior", "Priority", "Policy", "Health", "Inclusion", "pm25Avg")]

## surveys: drop items for which Inclusion == 0
surveys <- surveys[which(surveys$Inclusion == 1),]

## subset survey without Flashbarometer
surveys_sub <- surveys[which(surveys$source != "Flash Eurobarometer 360: Attitudes of Europeans towards air quality"),]

@@

## Visualizations by income group

```{r income group}

@@

# 1. Prepare topic-level data (shares within income group)
topics_data <- papers %>%
  pivot_longer(
    cols = c(Perceptions, Behavior, Priority, Policy, Health, None),
    names_to = "Topic",
    values_to = "Value") %>%
  filter(Value == 1, !is.na(Income.group)) %>%
  group_by(Topic, Income.group) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(Income.group) %>%
  mutate(share = n / sum(n)) %>%
  ungroup() %>%
  mutate(Income.group = factor(
    Income.group,
    levels = c("Low income", "Lower middle income", "Upper middle income", "High income")))

# 2. Prepare totals row data (absolute counts)
totals_data <- papers %>%
  filter(!is.na(Income.group)) %>%
  count(Income.group, name = "total")

@@

# 1. Prepare topic-level data (shares within income group)
topics_data_s <- surveys %>%
  pivot_longer(
    cols = c(Perceptions, Behavior, Priority, Policy, Health),
    names_to = "Topic",
    values_to = "Value") %>%
  filter(Value == 1, !is.na(Income.group)) %>%
  group_by(Topic, Income.group) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(Income.group) %>%
  mutate(share = n / sum(n)) %>%
  ungroup() %>%
  mutate(Income.group = factor(
    Income.group,
    levels = c("Low income", "Lower middle income", "Upper middle income", "High income")))

# 2. Prepare totals row data (absolute counts)
totals_data_s <- surveys %>%
  filter(!is.na(Income.group)) %>%
  count(Income.group, name = "total_items")

# 3. Define consistent factor levels for the y-axis
topic_levels_s <- c("Total items", unique(topics_data_s$Topic))

# 4. Plot with two independent fill scales
pdf("plots/Figure2_category_survey_by_income.pdf", width = 10)
ggplot() +
  # --- top totals row ---
  geom_tile(data = totals_data_s,
            aes(x = Income.group, y = "Total items", fill = total_items),
            color = "white", linewidth = 0.8) +
  geom_text(
    data = totals_data_s,
    aes(x = Income.group, y = "Total items",
        label = paste0("n=", total_items),
        color = ifelse(total_items > max(totals_data_s$total_items) * 0.4, "black", "black")), size = 4) +
  scale_fill_gradient(
    name = "Total items",
    low = "#F6DBA7", high = "#F7B32B",
    breaks = range(totals_data_s$total_items),         
  labels = range(totals_data_s$total_items)) +
  scale_color_identity() +

@@

## Visualizations by exposure group

```{r exposure group}

#-------------------------------------------------------------------------------
# get exposure groups for papers and surveys
#-------------------------------------------------------------------------------
papers$categoryPollution <- cut(papers$pm25Avg, breaks = c(0, 10, 15, 25, 35, 100), 
                                labels = c("Very low", "Low", "Medium", "High", "Very high"), 
                                right = FALSE)

surveys$categoryPollution <- cut(surveys$pm25Avg, breaks = c(0, 10, 15, 25, 35, 100), 
                                labels = c("Very low", "Low", "Medium", "High", "Very high"), 
                                right = FALSE)

# 1. Prepare topic-level data (shares within exposure group)
topics_data_poll <- papers %>%
  pivot_longer(
    cols = c(Perceptions, Behavior, Priority, Policy, Health, None),
    names_to = "Topic",
    values_to = "Value") %>%
  filter(Value == 1, !is.na(categoryPollution)) %>%
  group_by(Topic, categoryPollution) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(categoryPollution) %>%
  mutate(share = n / sum(n)) %>%
  ungroup() %>%
  mutate(categoryPollution = factor(
    categoryPollution,
    levels = c("Very low", "Low", "Medium", "High", "Very high")))

# 2. Prepare totals row data (absolute counts)
totals_data_poll <- papers %>%
  filter(!is.na(categoryPollution)) %>%
  count(categoryPollution, name = "total")

# 4. Plot with two independent fill scales
pdf("plots/Figure2_category_paper_by_exposure.pdf", width = 10)
ggplot() +
  # --- top totals row ---
  geom_tile(data = totals_data_poll,
            aes(x = categoryPollution, y = "Total papers", fill = total),
            color = "white", linewidth = 0.8) +
  geom_text(
    data = totals_data_poll,
    aes(x = categoryPollution, y = "Total papers",
        label = paste0("n=", total),
        color = ifelse(total > max(topics_data_poll$total) * 0.4, "black", "black")), size = 4) +
  scale_fill_gradient(
    name = "Total papers",
    low = "#F6DBA7", high = "#F7B32B",
    breaks = range(totals_data_poll$total),         
  labels = range(totals_data_poll$total)) +
  scale_color_identity() +

  # --- main topic rows ---
  geom_tile(data = topics_data_poll,
            aes(x = categoryPollution, y = Topic, fill = share),
            color = "white", linewidth = 0.8) +
  geom_text(
    data = topics_data_poll,
    aes(x = categoryPollution, y = Topic,
        label = scales::percent(share, accuracy = 1),
        color = ifelse(share > 0.25, "white", "black")),
    size = 4) +
  scale_fill_gradient(
    name = "Share within\nexposure group",
    low = "#f7f4f9", high = "#8B475D",
    labels = scales::percent,
    breaks = c(min(topics_data_poll$share), max(topics_data_poll$share))) +
  scale_color_identity() +
labs(x = "Exposure group", y = NULL) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 30, hjust = 1, size = 13),
    axis.text.y = element_text(size = 13),
    legend.position = "top")

# 1. Prepare topic-level data (shares within exposure group)
topics_data_poll_s <- surveys %>%
  pivot_longer(
    cols = c(Perceptions, Behavior, Priority, Policy, Health),
    names_to = "Topic",
    values_to = "Value") %>%
  filter(Value == 1, !is.na(categoryPollution)) %>%
  group_by(Topic, categoryPollution) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(categoryPollution) %>%
  mutate(share = n / sum(n)) %>%
  ungroup() %>%
  mutate(categoryPollution = factor(
    categoryPollution,
    levels = c("Very low", "Low", "Medium", "High", "Very high")))

# 2. Prepare totals row data (absolute counts)
totals_data_poll_s <- surveys %>%
  filter(!is.na(categoryPollution)) %>%
  count(categoryPollution, name = "total_items")

# 3. Define consistent factor levels for the y-axis
topic_levels_s <- c("Total items", unique(topics_data_s$Topic))

# 4. Plot with two independent fill scales
pdf("plots/Figure2_category_survey_by_exposure.pdf", width = 10)
ggplot() +
  # --- top totals row ---
  geom_tile(data = totals_data_poll_s,
            aes(x = categoryPollution, y = "Total items", fill = total_items),
            color = "white", linewidth = 0.8) +
  geom_text(
    data = totals_data_poll_s,
    aes(x = categoryPollution, y = "Total items",
        label = paste0("n=", total_items),
        color = ifelse(total_items > max(totals_data_poll_s$total_items) * 0.4, "black", "black")), size = 4) +
  scale_fill_gradient(
    name = "Total items",
    low = "#F6DBA7", high = "#F7B32B",
    breaks = range(totals_data_poll_s$total_items),         
  labels = range(totals_data_poll_s$total_items)) +
  scale_color_identity() +

  # --- main topic rows ---
  geom_tile(data = topics_data_poll_s,
            aes(x = categoryPollution, y = Topic, fill = share),
            color = "white", linewidth = 0.8) +
  geom_text(
    data = topics_data_poll_s,
    aes(x = categoryPollution, y = Topic,
        label = scales::percent(share, accuracy = 1),
        color = ifelse(share > 0.25, "white", "black")),
    size = 4) +
  scale_fill_gradient(
    name = "Share within\nexposure group",
    low = "#f7f4f9", high = "#8B475D",
    labels = scales::percent,
    breaks = c(min(topics_data_poll_s$share), max(topics_data_poll_s$share))) +
  scale_color_identity() +
labs(x = "Exposure group", y = NULL) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 30, hjust = 1, size = 13),
    axis.text.y = element_text(size = 13),
    legend.position = "top")
